Verification of Stencil Computations through Spatial Specifications

### 2016-07-07 paper.tex
conclusion edits

\begin{definition} The $\mathsf{toSpec}$ constructor is defined
in Figure~\ref{fig:to-spec}

\begin{figure}[t]
%\newcommand{\llower}{\vect{x}^L_i}
%\newcommand{\lupper}{\vect{x}^U_i}
\vspace{-0.7em}
\newcommand{\llower}{l}
\newcommand{\lupper}{u}
{\scalebox{0.97}{
\begin{minipage}{1\linewidth}
\begin{align*}
&\mathsf{toSpec} \; \llower \; \lupper = \\
&\hspace{-0.4em}\begin{cases}
%% FWD
\exact{\stenFwdS{\lupper}{i}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
\exact{\stenFwdSR{\lupper}{i}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
%% BWD
\exact{\stenBwdS{|\llower|}{i}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
\exact{\stenBwdSR{|\llower|}{i}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
%% CEN
\exact{\stenCenS{\lupper}{i}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.2em]
\exactR(\stenBwdS{|\llower|}{i} & \llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{i}) & \\[0.2em]
% BOUND
\upperB{\stenFwdS{\lupper}{i}} & \llower > 1 \\[0.2em]
\upperB{\stenBwdS{|\llower|}{i}} \quad & \llower < (-1)
 \\[0.2em]
% ABSOLUTE REP
\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
  \end{cases}
\end{align*}
\end{minipage}
}}
\caption{\textsf{toSpec} constructor of $\mathsf{A}(\specDNF)$ values.}
\label{fig:tospec}
\vspace{-0.9em}
\end{figure}

\noindent
The first two cases map to \texttt{forward} stencils since the lower
bound is $0$ or $1$, with the depth determined by the upper bound of
the span. If the lower bound is $1$, then the
$0$-offset point is missing so the region is given an
\texttt{irreflexive} attribute. The second two cases are similar for
\texttt{backward} but now the depth of the region is determined by the
absolute of the lower bound. In the fifth-case, where the lower bound
is negative and the upper bound is positive and they have the same
magnitude, then a \texttt{centered} region is constructed with
the depth as the common magnitude. If the magnitudes are not the same
(sixth case) then separated \texttt{forward} and \texttt{backward}
regions are summed, with their corresponding depths.

@@

One of the reasons this specification can catch many offset errors is that it is
a \texttt{readOnce} specification. This prevents offset errors leading to
duplicate access patterns be caught even if the resulting set of accesses are
part of the same family.
% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
% 42 cases were caught
% 6 cases were not caught
\section{Discussion}
\label{sec:discussion}

\paragraph{Program synthesis}

\anote{Any discussion left to do? - something on checking perhaps? access stencils?}

\subsection{Related work}

@@

\citet{kamil2016verified} propose a technique they call
\emph{verified lifting} for mapping low-level, possibly
greatly optimised, stencil computations into a high-level
predicate language using inductive program synthesis.
The aim is that the high-level abstract specifications can
be automatically translated into high-performance DSLs. Compared to our approach,
their specification language is more fine grained, capturing
the exact indexing pattern of a stencil. For example, the following
is a post-condition generated by their
compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
%
\begin{align*}
& \textit{post}(\texttt{a}, \texttt{b}) \equiv \forall \texttt{imin}+1
\leq \texttt{i} \leq \texttt{imax}, \texttt{jmin} \leq \texttt{j} \leq
\texttt{jmax} \\
& \qquad \texttt{a}(\texttt{i},\texttt{j}) =
\texttt{b}(\texttt{i}-1,\texttt{j}) + \texttt{b}(\texttt{i},\texttt{j})
\end{align*}
%
Notably, the bounds of the stencil and the exact indexing pattern
are captured. Our approach aims to be much more abstract, to
facilitate human reasoning, and avoid low-level lexical
errors. However, \textsc{stng} specifications are generated, \ie{},
are never written by a programmer, avoiding this
problem. Their inductive program synthesis approach allows
specifications to extracted even from optimised and unrolled code. For
example, the above post condition is generated from the code~\cite [p.3]{kamil2016verified}:
%
\begin{minted}{fortran}
do j=jmin, jmax
    t = b(imin, j)
    do i=imin+1,imax
        q = b(i, j)
        a(i, j) = q + t
        t = q
    end do
end do
\end{minted}
%
Thus, \textsc{stng} can understand cross-loop data-dependencies
in order to assign the above post-condition. CamFort is not currently
able to ascribe a specification in the above case as the use of
\texttt{t} on line 5 maps only to the definition on line 2, and not to
that on line 6. Extending our approach to follow loop-carried dependencies is
future work. An alternate approach for us would be to build on top of
the work of Kamil~\emph{et al.} by inferring our specifications from
\textsc{stng} (as an intermediate step).

% GPGPU
This work shares some similarities with efforts to verify kernels
written for General-Purpose Graphics Processing Unit (GPGPU)
programming, such as in \citet{Blom:2014:SoCP}. Stencils are a form of
kernel, and GPGPU programming can be viewed as a massively parallel
method of transforming a large matrix. However, that work focuses
mainly on the synchronisation of such kernels, and the avoidance of
data races, while we are interested in the correctness of stencils
embedded within a more typical general-purpose programming
language. Other work on GPGPU computation, such as
\citet{Zhang:2012:CGO}, has been primarily focused on generating
optimised code based upon on relatively simple specifications: to
provide performance while keeping the programmer's effort within
reason.

@@

% Pochoir
\citet{Tang:2011:SPAA} defines a specification language for writing
stencils embedded in C++ (with Cilk extensions) that are then compiled
into fast parallel programs based on trapezoidal decompositions with
hyperspace cuts. The Pochoir specification language is used for
describing the computing kernel, boundary conditions and shape of the
stencil. Programs are first compiled in C++ with a template library
that checks whether the annotations are used correctly, and then a
Pochoir preprocessor may be applied to generate the high-performance
parallel version that takes advantage of the Cilk compiler.  Pochoir
is aimed primarily at assisting programmers who may be reluctant to
implement the tricky cache-oblivious ``hypertrapezoidal'' algorithms
that are significantly faster than the typical looping implementations
of stencil computations. Like much of the related work, the main goal
is optimisation rather than catching bugs and verification of
program correctness.

By contrast, the work of \citet{Abe:2013:IPDPSW} has looked at the
correctness problem explicitly by bringing a form of model-checking to
verify certain stencil computations in partitioned global address
space languages. These are scenarios where an array is divided into
subarrays on multiple processors but each has the ability to access
the others' memory. The authors avoid the state explosion dilemma of
model-checking by relying upon the fact that accesses to the
``boundary elements'' will require a different method by virtue of the
non-locality of that memory. Since this access is performed in a
different manner, those can be identified automatically by static
analysis. To achieve this, the authors have designed a new language to
be used for writing stencil computations. Much of the specification
effort goes towards describing the nuts and bolts of distributing the
computation over multiple processors. The code for the stencil kernel
is generated from a relatively high-level specification. This approach
differs from CamFort because we are interested in integrating directly
into existing, legacy codebases and long-established languages such as
older versions of Fortran. We infer stencil specifications from
Fortran code and check annotations on stencil computations within
Fortran code.

There are many DSLs for stencil computation aimed at either
performance or verification, or both. Many eliminate out-of-bounds
errors, \eg{},~\citet{DBLP:journals/corr/abs-1109-0777}.

@@

In this paper we have described an annotation language which can be
used to specify the access pattern of a stencil computation. We
defined a denotational semantic model for our language and proved its
soundness. This gave a basis for an equational theory of
specifications and also justifies our inference and checking
procedure. We evaluated our approach by implementing the stencil inference and
checking procedure in CamFort, an open-source analysis tool\footnote{\url{https://github.com/camfort/camfort}} for
Fortran programs and running the inference process over a corpus of
1.1 million lines of Fortran code.

Our motivation for this work was to provide an intermediate
specification language to allow scientists to verify numerical
computations. The most common types of stencils we found consisted only of
reflexive region constants which are unlikely to give rise to
programming errors. However, around 10\% of stencils covered more
complicated program structures for which indexing errors are much more
plausible. We showed two examples of more complicated stencils for
which CamFort was able to detect a significant majority of possible
indexing errors.

Our definition of a stencil computation meant that we generated no
specification for a large number of iterated array accesses in the corpus. For
future work we will be looking at expanding the definition to allow us
to capture more of these examples.

>>>>

\begin{figure}[t]
%\newcommand{\llower}{\vect{x}^L_i}
%\newcommand{\lupper}{\vect{x}^U_i}
\vspace{-0.7em}
\newcommand{\llower}{l}
\newcommand{\lupper}{u}
{\scalebox{0.97}{
\begin{minipage}{1\linewidth}
\begin{align*}
&\mathsf{toSpec} \; \llower \; \lupper = \\
&\hspace{-0.4em}\begin{cases}
%% FWD
\exact{\stenFwdS{\lupper}{i}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
\exact{\stenFwdSR{\lupper}{i}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
%% BWD
\exact{\stenBwdS{|\llower|}{i}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
\exact{\stenBwdSR{|\llower|}{i}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
%% CEN
\exact{\stenCenS{\lupper}{i}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.2em]
\exactR(\stenBwdS{|\llower|}{i} & \llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{i}) & \\[0.2em]
% BOUND
\upperB{\stenFwdS{\lupper}{i}} & \llower > 1 \\[0.2em]
\upperB{\stenBwdS{|\llower|}{i}} \quad & \llower < (-1)
 \\[0.2em]
% ABSOLUTE REP
\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
  \end{cases}
\end{align*}
\end{minipage}
}}
\caption{\textsf{toSpec} constructor of $\mathsf{A}(\specDNF)$ values.}
\label{fig:to-spec}
\vspace{-0.8em}
\end{figure}

\begin{definition} The $\mathsf{toSpec}$ constructor is defined
in Figure~\ref{fig:to-spec}.
The first two cases map to \texttt{forward} stencils since the lower
bound is $0$ or $1$, with the depth determined by the upper bound of
the span. If the lower bound is $1$, then the
$0$-offset point is missing so the region is given an
\texttt{irreflexive} attribute. The second two cases are similar for
\texttt{backward} but now the depth of the region is determined by the
absolute of the lower bound. In the fifth-case, where the lower bound
is negative and the upper bound is positive and they have the same
magnitude, then a \texttt{centered} region is constructed with
the depth as the common magnitude. If the magnitudes are not the same
(sixth case) then separated \texttt{forward} and \texttt{backward}
regions are summed, with their corresponding depths.

@@

One of the reasons this specification can catch many offset errors is that it is
a \texttt{readOnce} specification. This prevents offset errors leading to
duplicate access patterns be caught even if the resulting set of accesses are
part of the same family.
% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
% 42 cases were caught
% 6 cases were not caught
\section{Related work}
\label{sec:discussion}

@@

\citet{kamil2016verified} propose a technique they call
\emph{verified lifting} for mapping low-level, possibly
greatly optimised, stencil computations into a high-level
predicate language using inductive program synthesis.
The aim is that the high-level abstract specifications can
be automatically translated into high-performance DSLs. Compared to our approach,
their specification language is more fine grained, capturing
the exact indexing pattern of a stencil. For example, the following
is a post-condition generated by their
compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
%
\begin{align*}
& \textit{post}(\texttt{a}, \texttt{b}) \equiv \forall \texttt{imin}+1
\leq \texttt{i} \leq \texttt{imax}, \texttt{jmin} \leq \texttt{j} \leq
\texttt{jmax} \\
& \qquad \texttt{a}(\texttt{i},\texttt{j}) =
\texttt{b}(\texttt{i}-1,\texttt{j}) + \texttt{b}(\texttt{i},\texttt{j})
\end{align*}
%
Notably, the bounds of the stencil and the exact indexing pattern
are captured. Our approach aims to be much more abstract, to
facilitate human reasoning, and avoid low-level lexical
errors. However, \textsc{stng} specifications are generated, \ie{},
are never written by a programmer, avoiding this
problem. Their inductive program synthesis approach allows
specifications to extracted even from optimised and unrolled code. For
example, the above post condition is generated from the code~\cite [p.3]{kamil2016verified}:
%
\begin{minted}{fortran}
do j=jmin, jmax
    t = b(imin, j)
    do i=imin+1,imax
        q = b(i, j);  a(i, j) = q + t
        t = q
    end do
end do
\end{minted}
%
Thus, \textsc{stng} can understand cross-loop data-dependencies
in order to assign the above post-condition. CamFort is not currently
able to ascribe a specification in the above case as the use of
\texttt{t} on line 5 maps only to the definition on line 2, and not to
that on line 6. Extending our approach to follow loop-carried dependencies is
future work. An alternate approach for us would be to build on top of
the work of Kamil~\emph{et al.} by inferring our specifications from
\textsc{stng} (as an intermediate step).

% GPGPU
Our work has some similarities with efforts to verify kernels
written for General-Purpose Graphics Processing Unit (GPGPU)
programming, such as in \citet{Blom:2014:SoCP}. Stencils are a form of
kernel, and GPGPU programming can be viewed as a massively parallel
method of transforming a large matrix. However, that work focuses
mainly on the synchronisation of such kernels and the avoidance of
data races, while we are interested in the correctness of stencils
embedded within a more typical general-purpose programming
language. Other work on GPGPU computation, such as
\citet{Zhang:2012:CGO}, has focused primarily on generating
optimised code based on relatively simple specifications: to
provide performance while keeping the programmer's effort within
reason.

@@

% Pochoir
\citet{Tang:2011:SPAA} define a specification language for writing
stencils embedded in C++ (with Cilk extensions) that are then compiled
into fast parallel programs based on trapezoidal decompositions with
hyperspace cuts. The Pochoir specification language is used for
describing the kernel, boundary conditions and shape of the
stencil. Programs are first compiled in C++ with a template library
that checks whether the annotations are used correctly, and then a
Pochoir preprocessor may be applied to generate the high-performance
parallel versions using Cilk.  Pochoir
is aimed primarily at assisting programmers who may be reluctant to
implement the high-performance tricky cache-oblivious
``hypertrapezoidal'' algorithms for stencils. 
%that are significantly faster than the typical looping implementations
%of stencil computations.
Like much of the related work, the main goal 
is optimisation rather than verifying program correctness.

By contrast, the work of \citet{Abe:2013:IPDPSW} has looked at the
correctness problem explicitly by bringing a form of model-checking to
verify certain stencil computations. This is in the context of
decomposing stencils for parallel computation on multiple processors
 in partitioned global address
space languages. %These are scenarios where an array is divided into
%subarrays on multiple processors but each has the ability to access
%the others' memory. The authors avoid the state explosion dilemma of
%model-checking by relying upon the fact that accesses to the
%``boundary elements'' will require a different method by virtue of the
%non-locality of that memory. Since this access is performed in a
%different manner, those can be identified automatically by static
%analysis. To achieve this, 
The authors have designed a new language to
be used for writing stencil computations. Much of the specification
effort goes towards describing the nuts and bolts of distributing the
computation over multiple processors. The code for the stencil kernel
is generated from a relatively high-level specification. This approach
differs from CamFort because we are interested in integrating directly
into existing, legacy codebases and long-established languages such as
older versions of Fortran. We infer stencil specifications from
Fortran code and check annotations on stencil computations within
Fortran code.

%There are many DSLs for stencil computation aimed at either
%performance or verification, or both. Many eliminate out-of-bounds
%errors, \eg{},~\citet{DBLP:journals/corr/abs-1109-0777}.

@@

In this paper, we described an abstract specification language 
for the access pattern of a stencil computation. Compared with the
related work (discussed above), our language is novel in that it does
not aim to replicate indexing in the specification, but instead
provides a set of abstract spatial combinators which capture a large number
of common patterns.

We evaluated our approach by implementing the stencil inference and
checking procedure ontop of CamFort, an open-source analysis
tool\footnote{\url{https://github.com/camfort/camfort}} for Fortran
programs and running the inference process over a corpus of 1.1
million lines of Fortran code.  Our motivation for this work was to
provide an intermediate specification language to allow scientists to
verify numerical computations.  Our evaluation showed that relatively
simple stencil comptuations are the norm, for which we can specify a
large proportion, comprising mainly reflexive region constants which
are less likely to give rise to programming errors.  However, around
10\% of stencils covered more complicated program structures for which
indexing errors are much more plausible. In the more detailed cases,
with larger specifications, the importance and impact of such
specifications grows.  We showed two examples of more complicated
stencils, for which CamFort was able to detect a significant majority
of possible indexing errors in our case studies.

Our definition of a stencil computation meant that we generated no
specification for a large number of iterated array accesses in the
corpus, such as stencils that incorporates reductions. For
future work, we will be looking into expanding the flexibility of 
our inference approach to more of these examples. Furthermore, our specification
language can be used more flexibly to give descriptions of data access
patterns in any context, not just within a stencil.
